Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-05 doc/JFLA2022/paper.tex
use macros for citing libraries

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}
\newcommand{\coq}{Coq\xspace}
\newcommand{\community}{Coq-community\xspace}
\newcommand{\gaia}{Gaia}
\newcommand{\alectr}{Alectryon}
\newcommand{\equations}{Equations}

@@

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library on discrete math, written for the \coq  proof assistant, and developed under the \community project. The proof scripts are
  accompanied with an electronic book, made with the help of the {\alectr} litterate proving tool.
  We present the evolution of the mathematical contents since
  former presentations at JFLA meetings.
  Then, we show how the structure of the project is determined   by two  requirements which must continuously satisfied:
   maintenance of the library with respect to several \community projects it requires and evolutions of \coq, and 
  consistency of the book with respect to the library. 
\end{abstract}

@@

\subsection{Vision}
The \emph{Hydra-battles} project, part of \community~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \community is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting \coq packages, and working collaboratively on documentation projects. The \emph{Hydra-battles} project demonstrates that these two goals are not independent: interesting \coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting \coq packages, we provide new documentation content, that contributes to the diversity of the thriving \coq documentation ecosystem.

@@

Futhermore, contrary to traditionally published books, the ``book'' that forms part of this project is intended to be forever evolving. As new \coq formalization patterns and proof techniques appear, the book can be adapted to demonstrate their use (in case they fit well with our applications).
%
By using modern maintenance techniques such as continuous integration and deployment, we can ensure that this documentation stays up to date with the latest \coq releases. With {\alectr}~\cite{alectryonpaper, alectryongithub}, we ensure that code and documentation are always in sync.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the \community project~\cite{CoqCommunity}, and documentation generated with {\alectr}~\cite{alectryonpaper, alectryongithub}.

@@

Computability is a key topic in Computer Science teaching. Moreover O'Connor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter to this formalization, with comments on the definitions and proofs, [counter]-examples and exercises, and  extract part of the Goedel library into a new sub-library of the Hydra-battles library: Ackermann.

@@

\subsection{Documentation with {\alectr}}

The hydra-battles book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the \coq development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}). The order in which these references appear in the book is independent of the structure of our libraries. Thus, we chose not to include our textbook as comments in the libraries (solution \emph{à la} Coqdoc).
\TODO[Pierre]{is the preceding sentence OK?}

@@

We solved this maintenance issue by moving to {\alectr}, a tool to automatically record \coq proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by {\alectr} from our \coq development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of \coq or versions of hydra-battles, we can easily spot unexpected changes).

@@

\begin{enumerate}
\item Mark the snippet in the \coq sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special {\alectr} annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of {\alectr}, so we had to extend it: the original {\alectr} did not have support for exporting to LaTeX, and it was towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended {\alectr} to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom {\alectr} \textit{driver}, a small (about 100 lines) Python program that leverages most of the {\alectr} toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as {\alectr}).

@@

\emph{Hydra-battles} wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the \coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

@@

We plan to extend our libraries in two main directions.
The Gaia project~\cite{Gaia}, also maintained on \community, contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \coq (SSreflect) of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \coq a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.

 \emph{Hydra-battles} is not limited to the study of ordinal numbers and applications.
 For instance, we are also developing 
a module about efficient exponentiation algorithms, and hope
to extend our project to new topics.

>>>>

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}
\newcommand{\coq}{Coq\xspace}
\newcommand{\community}{Coq-community\xspace}
\newcommand{\gaia}{Gaia\xspace}
\newcommand{\alectr}{Alectryon\xspace}
\newcommand{\equations}{Equations\xspace}
\newcommand{\Hydras}{Hydras \& Co.\xspace}

@@

\begin{abstract}
  \Hydras is a collaborative library on discrete math, written for the \coq  proof assistant, and developed under the \community project. The proof scripts are
  accompanied with an electronic book, made with the help of the \alectr litterate proving tool.
  We present the evolution of the mathematical contents since
  former presentations at JFLA meetings.
  Then, we show how the structure of the project is determined   by two  requirements which must continuously satisfied:
   maintenance of the library with respect to several \community projects it requires and evolutions of \coq, and 
  consistency of the book with respect to the library. 
\end{abstract}

@@

\subsection{Vision}
The \Hydras project, part of \community~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \community is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting \coq packages, and working collaboratively on documentation projects. The \Hydras project demonstrates that these two goals are not independent: interesting \coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the \gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting \coq packages, we provide new documentation content, that contributes to the diversity of the thriving \coq documentation ecosystem.

@@

Futhermore, contrary to traditionally published books, the ``book'' that forms part of this project is intended to be forever evolving. As new \coq formalization patterns and proof techniques appear, the book can be adapted to demonstrate their use (in case they fit well with our applications).
%
By using modern maintenance techniques such as continuous integration and deployment, we can ensure that this documentation stays up to date with the latest \coq releases. With \alectr~\cite{alectryonpaper, alectryongithub}, we ensure that code and documentation are always in sync.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the \community project~\cite{CoqCommunity}, and documentation generated with \alectr~\cite{alectryonpaper, alectryongithub}.

@@

Computability is a key topic in Computer Science teaching. Moreover O'Connor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter to this formalization, with comments on the definitions and proofs, [counter]-examples and exercises, and  extract part of the Goedel library into a new sub-library of the \Hydras library: Ackermann.

@@

\subsection{Documentation with \alectr}

The \Hydras book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the \coq development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}). The order in which these references appear in the book is independent of the structure of our libraries. Thus, we chose not to include our textbook as comments in the libraries (solution \emph{à la} Coqdoc).
\TODO[Pierre]{is the preceding sentence OK?}

@@

We solved this maintenance issue by moving to \alectr, a tool to automatically record \coq proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by \alectr from our \coq development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of \coq or versions of \Hydras, we can easily spot unexpected changes).

@@

\begin{enumerate}
\item Mark the snippet in the \coq sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special \alectr annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of \alectr, so we had to extend it: the original \alectr did not have support for exporting to LaTeX, and it was towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended {alectr to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom \alectr \textit{driver}, a small (about 100 lines) Python program that leverages most of the \alectr toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as \alectr).

@@

\Hydras wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the \coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

@@

We plan to extend our libraries in two main directions.
The \gaia project~\cite{Gaia}, also maintained on \community, contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \coq (SSreflect) of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \Hydras and the set-theoretic content of \gaia, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \coq a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.

 \Hydras is not limited to the study of ordinal numbers and applications.
 For instance, we are also developing 
a module about efficient exponentiation algorithms, and hope
to extend our project to new topics.
